\LARGE \bf Artifical Cognition for Social Human-Robot Interaction:\\ An Implementation

### 2016-06-14 paper.tex
Fixing language 3

\usepackage{xspace}
\newcommand{\ie}{{i.e.\xspace}}
\newcommand{\cf}{{\textit{cf\ }}}
\newcommand{\eg}{{e.g.\xspace}}
\newcommand{\al}{{et al.\xspace}}

@@

This knowledge model also largely excludes representation of continuous
phenomena (like time) or uncertain phenomena. When required (for instance for
action recognition), these are managed by dedicated components (like {\sc
Spark}), and are not exposed at the semantic level.

While alternatives like \emph{Answer Set Programming} have been successfully
investigated in robotics~\cite{Chen2010,Erdem2012}, in particular to deal with
non-monotonic reasoning, we did not actually encounter major reasoning issues
while working with OWL ontologies. We may reconsider this choice at a later
stage, but until now it has proven an effective framework to quickly explore
implementations of new cognitive abilities (for instance, it has been
conceptually and technically easy to add support for independent knowledge
models, one per agent the robot interacts with).

Anecdotally, and because ontologies and RDF statements remain conceptually simple
(compared to full logical languages like {\sc Prolog}), their adoption has also
effectively helped to grow awareness amongst colleagues on the significance of
the ``semantic level'' when developing new components for the robot.

@@

If a human informs the robot that a given object is indeed a bottle, the robot
derives more informations on the object. Conversely, if the human affirms that
``a car is a bottle'', the reasoner may detect logical contradictions (like
inconsistent sizes) and reject the assertion. The {\sc Dialogs} module relies on
such logical consistency checks when processing natural language inputs, both to
ensure that the verbal input has been correctly acquired and parsed, and also to
verify that what the human says is logically consistent.

We discuss the strengths and
weaknesses of this knowledge framework in Section~\ref{krs-discussion}.

@@

\emph{Grounding} (also called \emph{anchoring} when specifically referring to
the building of links between \emph{percepts} and \emph{physical
objects}~\cite{Coradeschi2003}) is the task consisting in building and
maintaining a bi-directional link between sub-symbolic representations (sensors
data, low-level actuation) and symbolic representations that can be manipulated
and reasoned about~\cite{Harnad1990}. This represents an important cognitive
skill, in particular in the human-robot interaction context: in that case, the
link that the robot has to established between its percepts and its symbol must
\emph{model} to a large extent the human representations in order to effectively
support communication.

@@

While the previous section provided an overview of our cognitive architecture
as well as its associated knowledge model, this section discusses its main
``building blocks''. Each of these components may have connections to several
others, and we invite the reader to refer to Figure~\ref{fig|archi} as a guide
if needed.

\begin{inparaenum} In our context, we call \emph{cognitive skills} the
deliberative behaviours that are \item \emph{stateful} (keeping track of
previous states is typically required for the component to perform adequately),
\item \emph{amodal} in that the skill is not inherently bound to a specific
perception of actuation modality, \item manipulate \emph{explicit semantics},
typically by the mean of symbolic reasoning, \item operate at the
    \emph{human-level}, \ie are legible to the humans, typically by acting at
similar levels of abstraction.\end{inparaenum}

We start by discussing first the main \emph{internal} cognitive capabilities,
implemented in the knowledge base itself, and then discuss successively the
situation assessment module {\sc Spark}, the dialogue processor {\sc Dialogs},
the symbolic task planner HATP, and finally the main features of our execution
controllers.

@@

As mentioned in the previous section, we use the Pellet open-source reasoner to
reason on the knowledge base. This enables several standard inference services:
\emph{consistency checking}, \emph{concept satisfiability},
\emph{classification} and \emph{realisation} (computation of the most specific
classes that an individual belongs to). In case of logical inconsistency, the
reasoner can also provide \emph{explanations} that are used in particular for
debugging.

Besides those, {\sc Oro} server implements several algorithms (presented
in~\cite{Ros2010b}) to identify
similarities and differences between concepts (classes or
instances): the \emph{Common Ancestors} algorithm, useful to
determine the most specific class(es) that include a given set of individuals;
the \emph{First Different Ancestors} algorithm that returns what can be
intuitively understood as the most generic types that \emph{differentiate} two
concepts; and \emph{clarification} and \emph{discrimination} algorithms that
play a key role in the process of \emph{interactive grounding} of the semantics
of concepts (we discuss this process at section~\ref{sect|com}). Clarification
and discrimination algorithms are based on what we call \emph{descriptors}, \ie
properties of individuals, either statically asserted in the common-sense
ontology, acquired by the robot through perception or pro-active questioning of
the human partner, or derived from other reasoning algorithms like the
\emph{Common Ancestors} and \emph{Different Ancestors}. The
\emph{discrimination} algorithm consists then in looking for discriminants, \ie
descriptors that allow a maximum discrimination among a set of individuals.

@@

Theory of Mind (originally defined in~\cite{Premack1978}) is the cognitive
ability that allows a subject to represent the mental state of another
agent, possibly including knowledge that contradicts the subject's own model: for
example, a book can be at the same time \emph{visible} for myself, and \emph{not
visible} for you. Children develop this skill, which is essential to understand others' perspectives during
interactions, around the age of three.

From a robotics point of view, it supposes the ability to build, store and
retrieve separate models of the beliefs of the agents the robot interacts with.
Our knowledge base implements such a mechanism: when the robots recognises that
a new agent has been created in the knowledge base, it initialises a new,
independent, ontology for this agent. All the ontologies that are created share
the same common-sense knowledge, but rely on each agent's perspective for the
actual instantiation: if the robot (geometrically) computes that the book is in its
own field of view, but not in the human one, the robot knowledge contains the
fact \stmt{book isVisible true} while the human model contains \stmt{book
isVisible false} (this computation, called \emph{perspective taking}, is
discussed in the next section).

One classical application of this cognitive skill is the so-called
\emph{False-Belief} experiment (also known as the \emph{Sally and Ann}
experiment)~\cite{Leslie2000}: a child is asked to watch a scene where two
people, A and B, manipulate objects. Then A leaves and B hides away one
object. When A comes back, we ask the child ``where do you think A will
look for the object?''. Before acquiring a theory of mind, children are not
able to separate their own (true) model of the world (where they know that
the object was hidden) from the model of A, which contains \emph{false
beliefs} on the world (A still thinks the object is at its original
position since he did not see B hiding it). Using separate knowledge models
in the knowledge base, we have been able to replicate this experience with
our robots~\cite{warnier2012when}.

@@

Memory has been studied at length in the cognitive psychology and
neuro-psychology communities: to give the main pointers, the idea of \emph{short-term} and
\emph{long-term} memory is due to Atkinson and Shiffrin~\cite{Atkinson1968};
Anderson~\cite{Anderson1976} proposes to split memory into \emph{declarative} (explicit)
and \emph{procedural} (implicit) memories; Tulving~\cite{Tulving1985} organises
the concepts of \emph{procedural}, \emph{semantic} and \emph{episodic} memories
into a hierarchy. Short-term memory is eventually refined with the concept of
\emph{working memory} by Baddeley~\cite{Baddeley2010}.
In the field of cognitive architectures, the {\sc Soar}
architecture~\cite{Lehman2006} is one of those that tries to reproduce a
human-like memory organisation. The GLAIR cognitive
architecture~\cite{Shapiro2009} also has a concept of long term/short term and
episodic/semantic memories.

It is worth emphasising that if memory is commonly associated to the process of
forgetting facts after a variable amount of \emph{time}, it actually covers
more mechanisms that are relevant to robotics, like selective remembering
triggered by a specific context or reinforcement learning.

The {\sc Oro} server features a mechanism to mimic only minimalistic forms of memory
families.  When new statements are inserted in the knowledge base, a
\emph{memory profile} is optionally attached to them.  Three such profiles are
predefined: {\it short term}, {\it episodic} and {\it long term}. They are
currently attached to different lifetime for the statements (respectively 10
seconds, 5 minutes and no time limit). After this duration, the statements are
automatically retracted.

@@

We rely however on this mechanism in certain cases: for instance, some modules
like the natural language processor use the {\it short term} memory profile to
mark for a few seconds important concepts that are currently manipulated by the
robot as \emph{active concepts}: if a human asks the robot ``Give
me all red objects'', the human, the \concept{Give} action, and each red
objects that are found are successively marked as \emph{active concepts} by
inserting statements such as \stmt{human type ActiveConcept} in the short-term
memory (which can be considered, in this case, to be a working memory). We use
this feature to trace certain knowledge-related processes. On the other hand,
our perception layer does not make use of this mechanism: as described in the
next section, the environment model of the robot is continuously updated and the
derived symbolic knowledge is therefore transient.

@@

    \node[anchor=north] at (-6.5,-5.3) {a) Agent-independent\\locations};
    \node[anchor=north] at (-2.1,-5.3) {b) Agent-dependent\\placements};
    \node[anchor=north] at (1.6,-5.3) {c) Visibility};
    \node[anchor=north] at (6.5,-5.3) {d) Reachability};

\end{tikzpicture}
}
    \caption{Functional overview of knowledge base ({\sc Oro} server, top part)
    and the geometric situation assessment module (\emph{{\sc Spark}}, bottom
    part). {\sc Spark} computes symbolic relationships between objects and
    agents, and feeds them to the knowledge base.}

@@

Anchoring perceptions in a symbolic model requires perception abilities and
their symbolic interpretation. We call \emph{physical situation assessment} the
cognitive skill that a robot exhibits when it assesses the nature and content of its
surroundings and efficiently monitors its evolution.

@@

We rely on a dedicated geometric and temporal reasoning module called {\sc
Spark} (\emph{SPAtial Reasoning \& Knowledge}, \cite{Sisbot2011}). It is a
situation assessment reasoner that generates symbolic knowledge from the
geometry of the environment with respect to relations between objects,
robots and humans (Figures~\ref{fig|spark-oro}
and~\ref{fig:sparkSubfigures}), also taking into account the different
perspective that each agent has on the environment.

{\sc Spark} embeds an \emph{amodal}~\cite{Mavridis2006} geometric model of the
environment that serves both as basis for the fusion of the perception
modalities and as bridge with the symbolic layer. This geometric model is made
of CAD 3D models of the objects, furnitures and robots, and full body, rigged
models of the humans, and it is updated at run-time by the robot based on its
sensors (in most of our studies, objects are identified and localised through 2D
fiducial markers, while humans are tracked with Kinect-like devices, optionally
assisted by motion capture to accurately track the head motion, which is
required to compute what the human is looking at).

@@

\begin{figure*}[ht!]
   \begin{center}
%
       \subfigure[Physical setting]{%
%           \label{fig:wuweiPhoto}
           \includegraphics[width=0.5\textwidth]{etat2-P1010769_brightened-v2.jpg}
       }%
       \subfigure[Corresponding 3D model view]{%
%          \label{fig:sparkScreenshot}
          \includegraphics[width=0.43\textwidth]{etat2_photo.png}
       }\\ %  ------- End of the first row ----------------------%
%
   \end{center}

>>>>

\usepackage{xspace}
\newcommand{\ie}{{i.e.\xspace}}
\newcommand{\cf}{{\textit{cf\ }}}
\newcommand{\eg}{{e.g.\xspace}}
\newcommand{\etal}{{et al.\xspace}}

@@

This knowledge model also largely excludes representation of continuous
phenomena (like time) or uncertain phenomena. When required (for instance for
action recognition), these are managed by dedicated components (like {\sc
Spark}, discussed at Section~\ref{sect|sit-ass}), and are not exposed at
the semantic level.

Alternative formalisms have been successfully investigated in robotics to
address some of these restrictions. \emph{Answer Set Programming} has been
used~\cite{Chen2010,Erdem2012} for instance to better supports non-monotonic
reasoning. We have also pointed in~\cite{lemaignan2015mutual} how modal logics like
the epistemic logics could be relevant to the particular field of social
human-robot interaction as they would allow for convenient representation of
alternative mental models. First-order logic and OWL ontologies have however
proved so far a simple, effective and sufficient symbolic framework for our
experimental applications.

Anecdotally, and because ontologies and RDF statements remain conceptually
simple (compared to full logical languages like {\sc Prolog} or modal logics),
their adoption has also effectively helped to grow awareness amongst colleagues
on the significance of the ``semantic level'' when developing new components
for the robot.

@@

If a human informs the robot that a given object is indeed a bottle, the robot
derives more informations on the object. Conversely, if the human affirms that
``a car is a bottle'', the reasoner may detect logical contradictions (like
inconsistent sizes) and reject the assertion. The {\sc Dialogs} module relies
on such logical consistency checks when processing natural language inputs,
both to ensure that the verbal input has been correctly acquired and parsed,
and also to verify that what the human says is logically consistent.  We
further discuss the strengths and weaknesses of this knowledge framework in
Section~\ref{krs-discussion}.

@@

\emph{Grounding} (also called \emph{anchoring} when specifically referring to
the building of links between \emph{percepts} and \emph{physical
objects}~\cite{Coradeschi2003}) is the task consisting in building and
maintaining a bi-directional link between sub-symbolic representations (sensors
data, low-level actuation) and symbolic representations that can be manipulated
and reasoned about~\cite{Harnad1990}. This represents an important cognitive
skill, in particular in the human-robot interaction context: in that case, the
link that the robot has to establish between percepts and symbols must
map as well as possible to the human representations in order to effectively
support communication.

@@

The previous section has provided an overview of our cognitive architecture
as well as the associated knowledge model. We discuss in this section its
building blocks. Each of these components have connections to several
others, and Figure~\ref{fig|archi} can be referred to as a guide.

\begin{inparaenum} We call \emph{cognitive skills} the
deliberative behaviours that are \item \emph{stateful}, \ie keeping track of
previous states is typically needed for the component to perform adequately;
\item \emph{amodal} in that the skill is not inherently bound to a specific
perception or actuation modality; \item manipulate \emph{explicit semantics},
typically by the mean of symbolic reasoning; \item operate at the
    \emph{human-level}, \ie are legible to the humans, typically by acting at
similar levels of abstraction.\end{inparaenum}

We present first the main \emph{internal} cognitive capabilities, implemented in
the {\sc Oro} knowledge base itself, and then discuss successively the situation
assessment module {\sc Spark}, the dialogue processor {\sc Dialogs}, the
symbolic task planner HATP, and finally the main features of our execution
controllers {\sc Shary} and {\sc pyRobots}. Note that greater details on each of
these modules can be found in their respective publications (the corresponding
references are provided hereafter).

@@

As mentioned in the previous section, we use the Pellet open-source reasoner to
reason on the knowledge base. It supports several standard inference processes:
consistency checking, concept satisfiability, classification and realisation
(computation of the most specific classes that a concept belongs to). In
case of logical inconsistency, the reasoner can also provide explanations (we
currently only use them for debugging purposes).

Besides, {\sc Oro} server implements several algorithms (presented
in~\cite{Ros2010b}) to identify
similarities and differences between concepts (classes or
instances): the \emph{Common Ancestors} algorithm, useful to
determine the most specific class(es) that include a given set of individuals;
the \emph{First Different Ancestors} algorithm that returns what can be
intuitively understood as the most generic types that \emph{differentiate} two
concepts; and \emph{clarification} and \emph{discrimination} algorithms that
play a key role in the process of \emph{interactive grounding} of the semantics
of concepts (we discuss this process at section~\ref{sect|com}). Clarification
and discrimination algorithms are based on what we call \emph{descriptors}, \ie
properties of individuals, either statically asserted in the common-sense
ontology, acquired by the robot through perception or pro-active questioning of
the human partner, or derived from other reasoning algorithms like the
\emph{Common Ancestors} and \emph{Different Ancestors}. The
\emph{discrimination} algorithm consists then in looking for discriminants, \ie
descriptors that allow a maximum discrimination among a set of individuals.

@@

Theory of Mind (originally defined in~\cite{Premack1978}) is the cognitive
ability that allows a subject to represent the mental state of another
agent, possibly including knowledge that contradicts the subject's own model: for
example, a book can be at the same time \emph{visible} for agent A, and \emph{not
visible} for agent B. Children develop this skill, which is essential to understand others' perspectives during
interactions, around the age of three.

From the point of view of robotics, it supposes the ability to build, store and
retrieve separate models of the beliefs of the agents the robot interacts with.
Our knowledge base implements such a mechanism: when the robots recognises that
a new agent has appeared, it initialises a new
independent ontology for this agent. All the ontologies that are created share
the same common-sense knowledge, but rely on each agent's perspective for the
actual instantiation: if the robot (geometrically) computes that the book is in its
own field of view, but not in the human one, the robot knowledge contains the
fact \stmt{book isVisible true} while the human model contains \stmt{book
isVisible false} (this computation, called \emph{perspective taking}, is
discussed in the next section).

One classical application of this cognitive skill is the so-called
\emph{False-Belief} experiment (also known as the \emph{Sally and Anne}
experiment, introduced by~\cite{wimmer1983beliefs} from an original
experimental setting by~\cite{baron1985does}): a child is asked to watch a scene where two
people, A and B, manipulate objects. At some point, A leaves and B hides away one
object. When A comes back, we ask the child ``where do you think A will
look for the object?''. Before acquiring a theory of mind, children are not
able to separate their own (true) model of the world (where they know that
the object was hidden) from the model of A, which contains \emph{false
beliefs} on the world (A still thinks the object is at its original
position since he did not see B hiding it). Using separate knowledge models
in the knowledge base, we have been able to replicate this experience with
our robots~\cite{warnier2012when}, in a manner similar to Breazeal
\etal\cite{breazeal2009embodied}.

@@

Memory has been studied at length in the cognitive psychology and
neuro-psychology communities: the idea of \emph{short-term} and \emph{long-term}
memory is due to Atkinson and Shiffrin~\cite{Atkinson1968};
Anderson~\cite{Anderson1976} proposes to split memory into \emph{declarative}
(explicit) and \emph{procedural} (implicit) memories; Tulving~\cite{Tulving1985}
organises the concepts of \emph{procedural}, \emph{semantic} and \emph{episodic}
memories into a hierarchy. Short-term memory is eventually refined with the
concept of \emph{working memory} by Baddeley~\cite{Baddeley2010}.  In the field
of cognitive architectures, the {\sc Soar} architecture~\cite{Lehman2006} is one
of those that tries to reproduce a human-like memory organisation. The GLAIR
cognitive architecture~\cite{Shapiro2009} also has a concept of long term/short
term and episodic/semantic memories.

It is worth emphasising that if memory is commonly associated to the process of
forgetting facts after a variable amount of \emph{time}, it actually covers
more mechanisms that are relevant to robotics, like priming
(concept pre-activation triggered by a specific
context~\cite{baxter2012modelling}) or reinforcement learning.

The {\sc Oro} server features a mechanism to mimic only minimalistic forms of memory
families.  When new statements are inserted in the knowledge base, a
\emph{memory profile} is attached to them.  Three such profiles are
predefined: {\it short term}, {\it episodic} and {\it long term}. They are
currently attached to different lifetime for the statements (respectively 10
seconds, 5 minutes and no time limit). After this duration, the statements are
automatically removed.

@@

We rely however on this mechanism in certain cases: some modules
like the natural language processor use the {\it short term} memory profile to
mark concepts that are currently manipulated by the
robot as \emph{active concepts}: if a human asks the robot ``Give
me all red objects'', the human, the \concept{Give} action, and each red
objects that are found are successively marked as \emph{active concepts} by
inserting statements such as \stmt{human type ActiveConcept} in the short-term
memory (which can be considered, in this case, to be a working memory). We use
this feature during dialogue disambiguation for instance. On the other hand,
our perception layer does not make use of this mechanism: as described in the
next section, the environment model of the robot is continuously updated and the
derived symbolic knowledge is therefore transient: it lasts only as long as the
environment remains in the same state.

@@

    \node[anchor=north] at (-6.5,-5.3) {a) Agent-independent\\(allocentric)
        spatial relations};
    \node[anchor=north] at (-2.1,-5.3) {b) Agent-dependent\\(egocentric) spatial
        relations};
    \node[anchor=north] at (1.6,-5.3) {c) Visibility/Pointing};
    \node[anchor=north] at (6.5,-5.3) {d) Reachability};

\end{tikzpicture}
}
    \caption{Functional overview of knowledge base ({\sc Oro} server, top part)
    and the geometric situation assessment module (\emph{{\sc Spark}}, bottom
    part). {\sc Spark} computes symbolic relationships between objects and
    agents, and exports them to the knowledge base.}

@@

Anchoring perceptions in a symbolic model requires perception abilities and
their symbolic interpretation. We call \emph{physical situation assessment} the
cognitive skill that a robot exhibits when it assesses the nature and content of its
surroundings and monitors its evolution.

@@

We rely on a dedicated geometric and temporal reasoning module called {\sc
Spark} (\emph{SPAtial Reasoning \& Knowledge}, presented in~\cite{Sisbot2011}).
It is a situation assessment reasoner that generates symbolic knowledge from the
geometry of the environment with respect to relations between objects, robots
and humans (Figures~\ref{fig|spark-oro} and~\ref{fig:sparkSubfigures}), also
taking into account the different perspective that each agent has on the
environment.  {\sc Spark} embeds an \emph{amodal} (as defined by Mavridis and
Roy in~\cite{Mavridis2006}: the different perceptual modalities are abstracted
away into a blended spatial model) geometric model of the environment that
serves both as basis for the fusion of the perception modalities and as bridge
with the symbolic layer. This geometric model is built from 3D CAD models of the
objects, furnitures and robots, and full body, rigged models of humans
(Figure~\ref{fig:sparkScreenshot}).  It is updated at run-time by the robot's
sensors (usually, a combination of vision-based tracking of 2D fiducial markers
to identify and localize objects, and Kinect-based skeleton tracking of humans,
optionally assisted by motion capture to accurately track the head motion, which
is required to compute what the human is looking at).

@@

\begin{figure*}[ht!]
   \begin{center}
%
       \subfigure[Physical setting]{%
           \label{fig:wuweiPhoto}
           \includegraphics[width=0.5\textwidth]{etat2-P1010769_brightened-v2.jpg}
       }%
       \subfigure[Corresponding 3D model view]{%
          \label{fig:sparkScreenshot}
          \includegraphics[width=0.43\textwidth]{etat2_photo.png}
       }\\ %  ------- End of the first row ----------------------%
%
   \end{center}
